---
title: "Distance based suicide Intervention programs, a multi-level Meta-Analysis and systematic review."
author: "Jim Schmeckenbecher, Katrin Rattner, Robert Cramer, Paul Plener, Anna Baran, Nestor Kapusta"
output: word_document
bibliography: 06_references.bib
csl: apa6.csl
after_body: appendix.md
header-includes:
  - \DeclareBibliographyCategory{asterisk}
  - \renewbibmacro*{begentry}{\ifcategory{asterisk}{\ensuremath{\ast}}{}}
  - \newcommand*{\nocitemeta}[1]{\nocite{#1}\addtocategory{asterisk}{#1}}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
op0 = function(x, d=0) sprintf(paste0("%1.",d,"f"), x) 
op = function(x, d=2) sprintf(paste0("%1.",d,"f"), x) 

```

# Abstract

Background: Distance based Interventions against suicidal thoughts and
behaviours are an increasingly found intervention type in the
literature, which has the potential to close the gap between needed and
provided care. Aims: Evaluate the effectiveness of distance based
interventions against suicidal thoughts and behaviours by implementing a
meta analysis and systematic review. Methods: We systematically searched
Web of Science, Scopus and Pubmed for all distance based
interventions;including all types outcome measures of suicidality. Data
was synthesised using a robust variance estimation corrected multilevel
meta-analysis . Results: 41 studies were included , reporting 121
outcomes . Effectiveness against suicidal thoughts was at low (*SMD =*
-0.17 CI95%[-0.21;-0.13]), but comparable to face-to-face interventions.
Against suicidal behaviours effectiveness were significantly lower, but
still significant (SMD = *SMD =* -0.06 CI95% [-0.08;-0.03] ),. Human
involvement had no significant impact on effectiveness. Conclusion:
Distance based interventions are an effective tool in a stepped care
approach aimed against suicidal thoughts, but for suicidal behaviours
more effective interventions are needed. Future research should focus on
the development of mass distributable autonomous programs against
suicidal ideation and plans.

# Introduction

Suicidal thoughts and behaviours are both a challenge for public health
and for service providers, given that annually 138 Million people
experience suicide ideation, 20.7 Million people attempt suicide
[@borges2010] and around 700.000 people die by suicide
[@worldhealthorganization2021]. Still only 17% to 56% of them receive
treatment [@bruffaerts2011]. Besides these undressed needs, low
treatment rates are linked to two structural barriers: Treatment cost
and availability [@bruffaerts2011].

Improving affordability and accessibility of treatment means to provide
suicide specific care in terms of tailored interventions according to
the patients stage of suicidal progression, rather than using a one fits
all solution. It has been suggested to implement a stepped care
approach, least restrictive care at early stages and to increase
restrictions gradually with advancement of suicidal progression
[@jobes2018].In this sense, easily available and affordable treatment
can lower treatment barriers and involve individuals otherwise
hesitating to seek help at early suicidal stages [@bruffaerts2011].
Early interventions at the stage of suicide ideation, have been
suggested to lower human suffering and to prevent future suicides
[@zuromski2019].

Distance based programs are least restrictive treatments, in terms of
local availability, affordability and service opening hours.
Under-serviced areas can be supported by both tele-health and Apps. A
one time purchase of an App, costs a fraction of a single therapy
session and is accessible 24h.

During the past two decades a number of randomized control trials
examining distance based programs have been published. Starting at the
turn of the millennium, with studies using phone-calls [@evans1999] and
post-cards [@motto2001], leading to crisis hotlines and e-mail
follow-ups [@luxton2020]. Recently the field has expanded to online
programs [@vanspijker2014; @franklin2016] and since the Covid-19
outbreak increasingly to tele-health approaches [@fernandez2021].
Several Meta-analyses have been published on subsets of distance based
programs [@milner2015; @torok2020].

To give recommendations for future research, our meta-analysis
differentiated between autonomous interventions (i.e. apps, online
programs) and human involved interventions (phone calls, postcards,
tele-health), which allows to investigate, whether the scalability of
autonomous interventions can be utilized without risking effectiveness.
To reach as many suicidal individuals as possible in early stages of
progression, distance based programs need to be scalable and
cost-effective. Autonomous interventions have superior scalability
compared to human involved interventions [@batterham2015], as they are
less expensive per intervention, less restricted by service opening
hours, they are translatable, and immediately available.

In order to draw practical conclusions we asked three questions,
implemented as moderation analyses: (a) Whether distance based programs
are effective against suicide ideation and/or against suicidal
behaviours, (b) How stable these interventions are over time and (c)
Whether effectiveness of such programs was independent from the chosen
control groups.

## Methods

The systematic search followed the Preferred Reporting Items for
Systematic Reviews and Meta-Analyses guidelines [@page2021] and was
Pre-Registered on Prospero under the pre-registration number:
CRD42020218791.

## Systematic Search

Search strings were defined using repeated searches combining the terms
suicide prevention OR intervention, with the intervention types, (e.g.)
Letter, App, Web-based, OR distance. The resulting search string was
tested and refined using two related meta-analyses, focusing on human
involved distance based interventions [@milner2015] and autonomous
interventions [@torok2020] (see Supplement A for the final strings).

Once search strings were established, the first one hundred search
results of Web of Science were examined together by the authors J.S. and
K.R., establishing a common degree of understanding. After which, both
authors independently searched Web of Science, Scopus and Pubmed;
Systematic searches were last updated on the ...... . Cohen's κ between
both authors was 0.806. [INTERNAL: As of 30.09 in WOS no new entries]

## Inclusion and exclusion criteria

All peer reviewed randomized control trial studies were included, which
investigated any form of distance based programs aimed at preventing
self-harming thoughts and/or behaviours, such as suicidal ideation,
suicidal planning, suicide attempts, non-suicidal self-injury (NSSI),
and suicide. Face-to-face meetings were allowed, if these were not part
of the intervention - i.e. for informing, testing or screening purposes.

Interventions which aimed at psychiatric disorders as a primary outcome
and combined outcome measures, such as the total score of SBQ-R, summing
thoughts and behaviours in a total score were excluded. Further, blended
programs, such as phone-calls including home visits were excluded.

## Data Extraction and Coding

Data was coded independently by two authors (J.S and K.R.). Where
possible non-imputed results, were coded. The following variables were
extracted : Author, Year, control group of study, Country of Study,
Sample type, sample size, intervention type, sex ratio, mean age, mean
age(SD), the outcome name (e.g. self-harm), intervention duration in
weeks, the participant attrition rate, the follow up time, standard mean
difference (*SMD)* and variance of *SMD*. In addition, all outcomes were
coded for the moderation analysis into subgroups (see Table 1).

TABLE 1.

+-----------+-----------------------------------------------------------+
| Moderator | **Outcome Name**                                          |
| group     |                                                           |
+===========+===========================================================+
| Acts      | Suicide, suicide attempts, self harming behaviours.       |
+-----------+-----------------------------------------------------------+
| Thoughts  | Suicidal thoughts, suicidal ideation, suicide plans.      |
+-----------+-----------------------------------------------------------+
| Human     | Phone calls, cognitive behavioural treatment,             |
| involved  | personalized letters or personalized e-mails.             |
+-----------+-----------------------------------------------------------+
| A         | Applications, websites, non-individualized letters or     |
| utonomous | non-individualized e-mails.                               |
+-----------+-----------------------------------------------------------+

: Table 1. Overview of outcome allocation to moderator group.

The authors compared finalized coding sheets, discussed differences and
re-coded affected studies until a unanimous result was achieved (see
Appendix).

## Risk of bias and publication bias

Risk of bias was assessed using the RoB-2 [@sterne2019] and Trim and
Fill was used as the publication bias detection
[@fernandez-castilla2021; @renkewitz2019].

## Statistical Method

To incorporate all outcomes of interest we used a three level
meta-analysis[@vandennoortgate2015][@cheung2019], with Robust variance
estimation (RVE) [@hedges2010; @moeyaert2017]. RVE return valid
confidence intervals in presence of dependent data [@park2019]. While
the three level model allowed for outcome-level heterogeneity
investigation [@vandennoortgate2013], RVE return valid confidence
intervals in presence of dependent data [@park2019]. Models were fitted
with restricted maximum likelihood estimation (REML), RVE correction was
based on Pustejovsky & Tipton [-@pustejovsky2016].

Calculations were done in R [@rcoreteam2020] using the package metafor
for the three level model [@viechtbauer2010c] and the package
clubSandwich [@pustejovsky2021a] for the RVE correction. R Code is
available on GitHub.

Calculations were done in R (4.1.1) [@rcoreteam2020] using the package
metafor (Version. X.XXX.) for the three level model [@viechtbauer2010c]
and the package clubSandwich (Version XXX) [@pustejovsky2021a] for the
RVE correction. R-Code is available as online supplement.

## Sensitivity Analysis

Given that NSSI [@americanpsychiatricassociation2013]and suicidal
behaviours [@joiner2005] may differ qualitatively, we employed two
sensitivity analyses: (a) Including Non-Suicidal Self-Injury (NSSI) as
an outcome, (b) Excluding suicide cases as an outcome. Deviation from
Pre-Registered Report

## Deviation from Pre-Registered Report

Given most recent developments in meta-analysis research we adapted the
procedure as follows: Firstly instead of the complex multi-level model ,
we employed a hybrid model of RVE correction and multilevel method
[@pustejovsky2021], thereby improving efficiency and reducing risk of
bias. Secondly, we followed the publication bias detection method
recommended by Fernández-Castilla et al. [-@fernandez-castilla2021],
although following the originally planned method [@renkewitz2019] would
generate the same results.

# Results

```{r include = FALSE}
source("04_Code.R", local = knitr::knit_global())

```

## Emerged Data

We identified 3,287 papers in the databases and additional 11 studies
from other sources (see Flow Chart), including `r max(Total$Study_ID)`
independent studies. The difference between included studies in the Flow
chart and reported independent studies is explained by follow-up studies
that use the same sample as their parent study. These follow-up studies
are reported as independent studies, but because they are statistically
dependent to their parent, they were treated as dependent within our
meta-analysis. Further one study included 3 statistically independent
RCT trials [@franklin2016].

INSERT FIGURE 1. Flow Chart

Overall, `r max(Total$Outcome_ID)` outcomes emerged, with
`r sum(Total$Thinking)` outcomes of the category thinking, predominantly
ideation and planning, and
`r max(Total$Outcome_ID) - sum(Total$Thinking)` outcomes of the category
acting, including studies examining mostly deliberate self-harm,
self-harm and suicide attempts.

`r max(Total$Outcome_ID) - sum(Total$Autonomous)` outcomes were found in
the human involved interventions category and `r sum(Total$Autonomous)`
outcomes were found for autonomous interventions. The median duration of
studies was `r op(median(Study_avarage$Duration_weeks))` weeks, with a
range of `r min(median(Total$Duration_weeks))` to
`r max(median(Total$Duration_weeks))` weeks. The median time between
post-intervention measures and follow up measures was
`r median(median(Study_avarage$Follow_up_Time, na.rm=TRUE))` weeks, with
a range of `r max(Total$Follow_up_Time,  na.rm = TRUE)` weeks to
`r min(Total$Follow_up_Time,  na.rm = TRUE)` weeks.

The median attrition rate was
`r op(median(Study_avarage$Total_Attrition, na.rm=TRUE))` % with a
maximum of `r max(Study_avarage$Total_Attrition, na.rm=TRUE)` % and a
minimum of 0 %.\

### Sample Characteristics

The studies included a total of `r Postn` at post-intervention
`r sum(Study_avarageFollow_up$n)` at follow up. Out of all,
`r op(mean(Study_avarage$Sex_Female))` % female and on average
`r op(mean(Study_avarage$Age_.mean.,na.rm=TRUE))` (SD=
`r op(mean(Study_avarage$Age_.SD., na.rm=TRUE))`) years old. In studies
the youngest reported mean sample age was
`r op(min(Study_avarage$Age_.mean.,na.rm=TRUE))` (SD= 1.4) years, the
oldest mean sample age was
`r op(max(Study_avarage$Age_.mean.,na.rm=TRUE))` (SD= 11.3)years.

Out of a total of n= `r max(Total$Study_ID)` studies, most data was
retrieved from the United States (*k=* 10), followed by Australia (*k* =
8); and from non-westernised educated industrialised democracies (WEIRD;
*k =* 6) countries.

## Main Analysis

Distance based interventions were effective against suicidal thoughts
and behaviours, standardized mean difference (SMD) =
`r op(OverallRVECI$beta)` CI95%[`r op(OverallRVECI$CI_L)`;
`r op(OverallRVECI$CI_U)`]; Heterogeneity was significant at *Q* (df =
`r Overall$dfs`)= `r op(Overall$QE)`, *p =* `r op(Overall$QEp)`.

The moderation analysis was significant, showing that distance based
programs were (*SMD=* `r op(Overall.TRVECI$beta[2])`
CI95%[`r op(Overall.TRVECI$CI_L[2])`; `r op(Overall.TRVECI$CI_U[2])`])
more effective against suicidal thoughts than suicidal behaviours. The
average effectiveness against suicidal thoughts was *SMD=*
`r op(Overall.T2RVECI$beta[1])` CI95%[`r op(Overall.T2RVECI$CI_L[1])`;
`r op(Overall.T2RVECI$CI_U[1])`] compared to *SMD=*
`r op(Overall.TRVECI$beta[1])` CI95%[`r op(Overall.TRVECI$CI_L[1])`;
`r op(Overall.TRVECI$CI_U[1])`] for suicidal behaviours. Heterogeneity
was non-significant *Q* (df = `r Overall.T$dfs`) = `r op(Overall.T$QE)`,
*p =* `r op(Overall.T$QEp)`. See Appendix C for study average effects
against suicidal thoughts and suicidal acts.

Given small study numbers, the comparison of waitlist and attention
placebo groups showed statistically not trustworthy results according to
the profile likelihood plots.

Therefore, waitlist and attention placebo were combined into a combined
control group, and compared to TAU.

Comparing the combined control-group vs. TAU, TAU groups were
significantly less impacted by distance based interventions (S*MD=*
`r op(Overall.CRVECI$beta[1])` CI95%[`r op(Overall.CRVECI$CI_L[1])`;
`r op(Overall.CRVECI$CI_U[1])`]), than studies using a waitlist or
attention placebo designs. Heterogeneity remained significant = *Q* (df
= `r OverallC$dfs`) = `r op(OverallC$QE)`, *p =* `r op(OverallC$QEp)`.

Covariance was investigated by simultaneous visualization of
control-group and outcome category. Suicidal behaviours and suicidal
thoughts were unevenly distributed between different control-groups (see
figure 2).

An exploratory analysis including both moderators was implemented.

When including both moderators, the difference between TAU and the
combined control group was non- significant, with *SMD=*
`r op(Overall.TRVECITT$beta[3])` CI95%[`r op(Overall.TRVECITT$CI_L[3])`;
`r op(Overall.TRVECITT$CI_U[3])`], but suicidal acts and suicidal
thoughts remained a significant moderator, with *SMD =*
`r op(Overall.TRVECITT$beta[2])` CI95%[`r op(Overall.TRVECITT$CI_L[2])`;
`r op(Overall.TRVECITT$CI_U[2])`] in favour of suicidal thoughts;
heterogeneity was not significant = *Q* (df = `r Overall.TT$dfs`) =
`r op(Overall.TT$QE)`, *p =* `r op(Overall.TT$QEp)`.

INSERT: FIGURE 2. Funnel Plot

Effectiveness of distance based interventions decreased between
time-points at *SMD =* `r op(Overall.FCRVECI$beta[2])`
CI95%[`r op(Overall.FCRVECI$CI_L[2])`; `r op(Overall.FCRVECI$CI_U[2])`]
non-significantly. Heterogeneity was significant = *Q* (df =
`r OverallF$dfs`) = `r op(OverallF$QE)`, *p =* `r op(OverallF$QEp)`.

Human involvement had with *SMD=* `r op(Overall.IRVECI$beta[2])`
CI95%[`r op(Overall.IRVECI$CI_L[2])`; `r op(Overall.IRVECI$CI_U[2])`] an
non-significant negative impact on effectiveness. Heterogeneity was
significant = *Q* (df = `r Overall.I$dfs`) = `r op(Overall.I$QE)`, *p =*
`r op(Overall.I$QEp)`.

## Sensitivity Analysis

*Both, inclusion of NSSI (k = 6) and exclusion of suicide studies (k= 4)
had negligible impacts on the subgroup of behaviour outcomes. Inclusion
of NSSI increased the effectiveness SMD=* `r op(Overall.TRVE2$beta[1])`
to *SMD*= `r op(Overall.TRVE$beta[1])`. and *xclusion of suicide studies
decreased effectiveness from SMD =* `r op(Overall.TRVE$beta[1])` to
*SMD*= `r op(Overall.TSRVE$beta[1])`.

## Publication Bias and Risk of Bias Assessment

Risk of bias of all *independent* studies was mixed (see Figure 3.).
Using Trim and Fill, no publication bias could be observed.

INSERT: FIGURE 3. Rob-2 Study Quality

# Discussion

In this meta-analysis we investigated the effectiveness of distance
based programs in reducing suicidal thoughts and behaviours. On average
distance based programs reduced both suicidal thoughts and behaviours.
The quality of evidence was good, given a considerable number of RCTs,
with a notable number of high and medium quality studies and no observed
publication bias.

### Contextualising results with other Meta-Analyses

We showed that suicidal behaviours were reduced
(`r op(Overall.T2RVECI$beta[1])` CI95%[`r op(Overall.T2RVECI$CI_L[1])`;
`r op(Overall.T2RVECI$CI_U[1])`]) by distance based programs. These
reductions were in line with face-to-face therapeutic interventions. One
recent meta-analysis on face-to-face therapeutic interventions
[@ougrin2015] reported a reduction of *SMD =* -0.11 CI95% [-0.38; 0.15]
for suicide attempts and a current meta-analysis [@kothgassner2020]
reported a reduction of SMD = -0.13 CI95% [-0.22;-0.04] for self harm.
In contrast to these short-term interventions, a meta-analysis of
higher-frequency and long-term psychoanalytic face-to-face interventions
[@briggs2019] reported stronger results, for suicide attempts *SMD =*
-0.24 CI95%[−0.50 to 0.03] and for self harm between *SMD*= -0.73 CI95%
[-1.22;-0.22] and *SMD*= -0.15 CI95% [-3.88;0.09]; depending on how self
harm was measured. Given that most distance based programs had very low
frequencies (e.g. letter sent every few months), while Ougrin et al.
[-@ougrin2015] and Kothgassner et al. [-@kothgassner2020] mostly
included low frequency face-to-face interventions and Briggs et al.
[-@briggs2019] mostly included higher frequency interventions. Frequency
of intervention seems to be a moderating factor for effectiveness.

Although distance-based interventions seem to have a small effect size
in comparison to face-to face interventions, especially high-frequency
long-term interventions, they might have a role in the prevention of
suicidal behaviour due to their availability, scalability and
cost-efficacy. Even more, it remains to be analysed, whether the higher
effectiveness of face-to-face interventions, which is based on rather
small samples, would remain superior when their power would be
increased. For example, although the effect size of the meta-analysis by
Briggs et al. [-@briggs2019] seems four times higher than the effect
size of distance based interventions, our effect is more precise and
statistically significant.

When we also draw on the effect-size estimates and their confidence
intervals, we can observe the clinical potential of an intervention. For
example, the effect size reported by Briggs et al. (2019) against
suicide attempts, expressed as Number Needed to Treat, (NNT) is = 7.4
CI95% [Inf; 3.6], means that the interventions may be ineffective
(Infinite number needed), but it could help up to every fourth person.
In contrast, our results of distance based programs, expressed as *NNT
=* 29.5 CI95% [59; 22.1] means that a reduction will occur with 95%
certainty, but *at best* only every twenty-second person will be helped
and on average one of 30 interventions will be helpful. Given the
seriousness of suicidal behaviours and the currently low effectiveness
of interventions against it, more research on potentially more effective
treatment is needed. For example, we suggest more research into high
frequency interventions and on the question if the trend of larger
effect-sizes in psychoanalytical therapies [@briggs2019] and DBT-A
therapies [@kothgassner2020] is retained, given more data. However,
despite the low effect of distance based programs on suicidal behaviour,
they nevertheless might have a role in the spectrum of preventive
approaches. Especially the fact of high barriers to treatment seeking,
especially to face-to-face treatment, are an argument to disseminate
distance based interventions as a means for involvement where other
treatment is not available or not accepted by those in need.

### 

### Effectiveness of distance based programs for suicidal thoughts 

Similarly to suicidal behaviours, suicidal thoughts were reduced by
distance based programs (`r op(Overall.T2RVECI$beta[1])`
CI95%[`r op(Overall.T2RVECI$CI_L[1])`;
`r op(Overall.T2RVECI$CI_U[1])`]). The result is line with face-to-face
interventions according to Kothgassner et al. [-@kothgassner2020] (SMD =
-0.31 CI95% [-0.5; -0.12]). Comparing the intervention-type subgroups in
Kothgassner et al. [-@kothgassner2020], to our results, some
face-to-face interventions like group therapy (SMD = 0.01
CI95%[-0.17;0.18]) and Brief Interventions (SMD = -0.12CI95%[-0.27;0.03]
tended to be less effective, with other forms of intervention like DBT-A
(-0.48 CI95%[-0.8;-0.17] tending to be more effective; but none differed
significantly.

Distance based programs can be applied on multiple levels of the stepped
care model for suicide care [@jobes2018]. This model includes 5 levels
of intervention, ranging from least restrictive Telephone (level 1),
Brief interventions (2), outpatient care (3), partial hospitalisation
(4) to most restrictive interventions like inpatient care/full
hospitalisation(5).

As such both the least costly and restrictive level of stepped care,
'telephone interventions and follow ups' and the second least costly and
restrictive level, 'Brief interventions and follow ups' can be
supplemented or supplanted by distance based interventions. Further,
given that effect-sizes of distance based interventions were similar to
other face-to-face interventions, even the level 3- outpatient care, can
be supplemented with distance based programs. In summary, distance based
interventions are capable to supply evidence-based, least-restrictive,
and cost effective help. Distance based programs can even be a
substitute to outpatient care, where it is not immediately available or
not used by individuals seeking help.

### Research recommendations

According to our results, autonomous distance based interventions (ADBI)
(SMD= `r op(Overall.I2RVECI$beta[1])`
CI95%[`r op(Overall.I2RVECI$CI_L[1])`;
`r op(Overall.I2RVECI$CI_U[1])`]), were as effective as human involved
interventions (SMD= `r op(Overall.IRVECI$beta[1])`
CI95%[`r op(Overall.IRVECI$CI_L[1])`; `r op(Overall.IRVECI$CI_U[1])`]).
But ADBI promise a better scalability [@batterham2015], making them more
feasible for studies with larger sample sizes and replication studies.

Studies with large sample sizes or study sets using close replications
are needed, as only these allow us to control for the characteristics of
distance based interventions and by such, to understand which components
are most effective.

In general, the assumptions about Distance based interventions (DBI) for
mental health often lack evidence [@musiat2014]. For example, some
stated advantages of DBIs in suicide prevention are still unproven
assumptions, for example that 24h availability of such an intervention
might be advantageous. The best proven assumptions derived from
meta-analytical evidence are delivered for cost effectiveness,
acceptability and satisfaction of DBIs. However, this evidence is
available for mental health interventions in general [@eze2020;
@musiat2014], but not for the primary endpoint of suicidal behaviour or
thoughts.

In this sense, the development of well powered ADBIs for suicide
prevention is still is in its infancy and needs to addressed in future
more scrupulously.

### Implications for clinical practice

Suicide research shows that all interventions, face-to-face-,
Human-Involved Distance Based- as well as autonomous Interventions are
effective in reducing suicide. Further, all of these interventions were
more effective against suicidal thoughts than against suicidal acts.
Therefore, there is a strong argument for the implementation of such
interventions among individuals with suicide ideation, as recently
discussed by Joiner and Jobs[-@jobes2019] . The availability of
face-to-face interventions in mental health clinics and among individual
mental health professionals such as psychiatrists and psychotherapists
is unevenly distributed geographically [@kapusta2010;@pirkola2009] and
may be limited in pandemic containment efforts against Covid-19. The
resulting unavailability may be compensated for by ADBIs, which are not
affected by geographic distribution or lockdowns.In addition the reality
of psychiatric treatment includes high costs for individuals or the
public health system [@wittchen2011a], depending on whether psychiatric
treatment is covered by insurance. This often results in long waiting
times for patients in need [@zepf2003]. In both cases, ADBIs can help
mitigate the negative effects of barriers to help seeking, by offering
an intermediate alternative, thereby bridging waiting times and at lower
costs. Finally ADBI can help to scale up mental health services
especially in countries with limited mental health expenditures and low
resources, such as low and middle income countries (LMIC), as called by
Chisholm et al. [-@chisholm2007] and thereby contribute to the
Sustainable Mental Health Development Goals, set out be the UN
[@patel2018].

## Limitations

Given our comprehensive approach, some limitations should be noted.

Firstly we did not include any grey literature. This can cause selection
bias, as the publication bias suppresses non-significant results. But,
as RCT studies in a multidisciplinary field (clinical psychologists, IT
experts, physicians...) are very costly, expecting grey literature
research is rather unreasonable. And non-publication due to reporting
non-significant results is also unlikely, given that almost all studies
included non-significant results. This decision reduced work load
notably and is in line with previous meta-analyses to this field
[@torok2020; @milner2015 ].

The second potential limitation is that most studies included in this
meta-analysis are already covered by different previously published
meta-analyses [@torok2020; @milner2015 ]. However, these previously
published meta-analyses used a method - Hedges-Olkin meta-analysis -
which can only include independent data. In contrast our method
(Multi-Level Meta-Analysis) allows inclusion of non-independent data, by
this all relevant data from a study can be utilized [@cheung2019].

Utilizing all relevant data gives multiple advantages, foremost higher
precision and less bias risk (A) Precision and non-independent data,
including multiple non independent data points per study increases
precision. Further it allows important moderator analysis to be
implemented in one model, which allows to weigh evidence according to
its information value. In contrast previous meta-analysis had to use
independent subgroup analysis, which lowers precision notably and only
allows for an indirect comparisons of effectiveness. (B) Bias risk, our
meta-analysis is the first to include all relevant outcomes. In contrast
previous Hedges-Olkin meta-analyses [@torok2020; @milner2015 ] were
forced to select one representative outcome per independent analysis.
Due to this potential of a selection bias can therefore not be excluded
in these older meta-analyses. Based on these points and the fact that we
updated and broaden the systematic searches of previous meta-analysis,
the current meta-analysis substantially adds to the research field.

But while a multilevel meta-analysis gives all the above stated
advantages, it also produces potential limitations. First The employed
method requires more studies to reach adequate power, than hedges Olkin
meta-analysis. And the amount of included studies was low, in turn the
analysis might have been underpowered [@tanner-smith2014]. But results
were reliable, based on Profile Likelihood Plots of metafor and the
reported degrees of freedom reported by the RVE Correction
[@pustejovsky2016]. Second, RVE corrected models do not report
heterogeneity estimations, therefore heterogeneity estimations of the
MLM were reported. However *Q-* test results are not biased by
dependency and therefore statistically valid, while power of the
*Q*-Test was sufficient [@maeda2016].

Summarizing: Distance based interventions, particularly autonomous
distance based interventions are an effective and cheap way of
supporting treatment, specially against suicidal thoughts and in
situations where availability of face-to-face treatments are limited.

nocite:\|
@amadeo2015,@batterham2018,@beautrais2010,@bertolote2010,@bush2017,@carter2005,@carter2007,@carter2013,@cedereke2002,@christensen2013,@comtois2019,@dejaegere2019,@evans2005,@evans1999,@franklin2016,@guille2015,@hassanian-moghaddam2011,@hassanian-moghaddam2017,@hetrick2017,@hill2019,@hooley2018,@kasckow2016,@luxton2020,@motto2001,@mouaffak2015,@mousavi2014,@niederkrotenthaler2020,@oconnor2017,@robinson2012,@rodante2020,@tighe2017,@vaiva2018,@vaiva2006,@vanspijker2014,@vanspijker2018,@wilks2018,@wilksch2019,@wei2013

# References

::: {#refs}
:::

# Appendix

## Online Supplement 1

### Search Strings:

Note: Search string was adapted in Scopus adding: TITLE-ABS-KEY. This is
implied when using WOS and Pubmed. As in WOS and Pubmed Title, Abstract
and Keywords were also searched, both strings are in practice identical.

#### For WOS and Pubmed

(("Suic\*" OR "Suicide prevention" OR "self harm\*" OR "self
poisoning\*" OR "self injur\*" OR "self mutilation" ) AND
("telehealth"OR"postcard\*"OR "onlin\*"OR"Online Intervention" OR
"Online Prevent\*" OR "E Intervention" OR "E-Intervention" OR "E
Prevention" OR "Electronic Intervention" OR "Electronic Prevention" OR
"Mobile Intervention" OR "Mobile Prevention" OR "Web-Based\*" OR "Web
Based\*"  OR "Online Support" OR "E Therapy" OR "e-mail\*" OR "e mail\*"
OR "App" OR "Apps" OR "App-Assis\*" OR "mobile-App" OR "mobile health
intervention" OR "telephone" OR "phone based" OR "letter\*") AND (RCT OR
Random\*))

#### Scopus:

( TITLE-ABS-KEY ( ( "Suic\*"  OR  "Suicide prevention"  OR  "self
harm\*"  OR  "self poisoning\*"  OR  "self injur\*"  OR  "self
mutilation" )  AND  ( "telehealth"  OR  "postcard\*"  OR  "onlin\*"  OR 
"Online Intervention"  OR  "Online Prevent\*"  OR  "E Intervention"  OR 
"E-Intervention"  OR  "E Prevention"  OR  "Electronic Intervention"  OR 
"Electronic Prevention"  OR  "Mobile Intervention"  OR  "Mobile
Prevention"  OR  "Web-Based\*"  OR  "Web Based\*"  OR  "Online Support" 
OR  "E Therapy"  OR  "e-mail\*"  OR  "e mail\*"  OR  "App"  OR  "Apps" 
OR  "App-Assis\*"  OR  "mobile-App"  OR  "mobile health intervention" 
OR  "telephone"  OR  "phone based"  OR  "letter\*" )  AND  ( rct  OR 
random\* ) ) )

## Online Supplement 2

### Code.R

Coding_Sheet.csv
