---
title: "Distance based suicide Intervention programs, a multi-level Meta-Analysis and systematic review."
output: word_document
bibliography: references.bib
csl: apa6.csl
after_body: appendix.md
header-includes:
  - \DeclareBibliographyCategory{asterisk}
  - \renewbibmacro*{begentry}{\ifcategory{asterisk}{\ensuremath{\ast}}{}}
  - \newcommand*{\nocitemeta}[1]{\nocite{#1}\addtocategory{asterisk}{#1}}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
op0 = function(x, d=0) sprintf(paste0("%1.",d,"f"), x) 
op = function(x, d=2) sprintf(paste0("%1.",d,"f"), x) 

```

# Abstract

Background: Distance based Interventions against suicidal thoughts and
behaviours are an increasingly found intervention type in the
literature,which has the potential to close the gap between needed and
provided care. Aims: Evaluate the effectiveness of distance based
interventions against suicidal thoughts and behaviours by implementing a
meta analysis and systematic review. Methods: We systematically searched
Web of Science, Scopus and Pubmed for all distance based interventions;
including all types outcome measures of suicidality. Data was
synthesised using a robust variance estimation corrected multilevel
meta-analysis . Results: 41? studies were included , reporting 121
outcomes . Effectiveness against suicidal thoughts was at low (SMD= *SMD
=*-0.17 CI95%[-0.21;-0.13]), but comparable to face-to-face
interventions . Against suicidal behaviours effectiveness were
significantly lower, but still significant (SMD*=* -0.06 CI95%
[-0.08;-0.03] ). Human involvement had no significant impact on
effectiveness. Conclusion: Distance based interventions are an effective
tool in a stepped care approach aimed against suicidal thoughts, but for
suicidal behaviours more effective interventions are needed. Future
research should focus on the development of mass distributable
autonomous programs against suicidal ideation and plans.

# Introduction

Suicidal thoughts and behaviours are both a challenge for public health
and for service providers, given that annually 138 Million people
experience suicide ideation, 20.7 Million people attempt suicide
[@borges2010a] and around 700.000 people die by suicide
[@worldhealthorganization2021]. Still only 17% to 56% of them receive
treatment [@bruffaerts2011]. Besides these unaddressed needs, low
treatment rates are linked to two structural barriers: Treatment cost
and availability [@bruffaerts2011].

Improving affordability and accessibility of treatment means to provide
suicide specific care in terms of tailored interventions according to
the patients stage of suicidal progression, rather than using a one fits
all solution. It has been suggested to implement a stepped care
approach, least restrictive care at early stages and to increase
restrictions gradually with advancement of suicidal progression
[@jobes2018a].In this sense, easily available and affordable treatment
can lower treatment barriers and involve individuals otherwise
hesitating to seek help at early suicidal stages [@bruffaerts2011].
Early interventions at the stage of suicide ideation, have been
suggested to lower human suffering and to prevent future suicides
[@zuromski2019].

Distance based programs are least restrictive treatments, in terms of
local availability, affordability and service opening hours.
Under-serviced areas can be supported by both tele-health and Apps. A
one time purchase of an App, costs a fraction of a single therapy
session and is accessible 24h.

During the past two decades a number of randomized control trials
examining distance based programs have been published. Starting at the
turn of the millennium, with studies using phone-calls [@evans1999] and
post-cards [@motto2001], leading to crisis hotlines and e-mail
follow-ups [@luxton2020]. Recently the field has expanded to online
programs [@vanspijker2014; @franklin2016] and since the Covid-19
outbreak increasingly to tele-health approaches [@fernandez2021].
Several Meta-analyses have been published on subsets of distance based
programs [@milner2015; @torok2020].

To give recommendations for future research, our meta-analysis
differentiated between autonomous interventions (i.e. apps, online
programs) and human involved interventions (phone calls, postcards,
tele-health), which allows to investigate, whether the scalability of
autonomous interventions can be utilized without risking effectiveness.
To reach as many suicidal individuals as possible in early stages of
progression, distance based programs need to be scalable and
cost-effective. Autonomous interventions have superior scalability
compared to human involved interventions [@batterham2015], as they are
less expensive per intervention, less restricted by service opening
hours, they are translatable, and immediately available.

In order to draw practical conclusions we asked three questions,
implemented as moderation analyses: (a) Whether distance based programs
are effective against suicide ideation and/or against suicidal
behaviours, (b) How stable these interventions are over time and (c)
Whether effectiveness of such programs was independent from the chosen
control groups.

## Methods

The systematic search followed the Preferred Reporting Items for
Systematic Reviews and Meta-Analyses guidelines [@page2021] and was
Pre-Registered on Prospero under the pre-registration number:
CRD42020218791.

## Systematic Search

Search strings were defined using repeated searches combining the terms
suicide prevention OR intervention, with the intervention types, (e.g.)
Letter, App, Web-based, OR distance. The resulting search string was
tested and refined using two related meta-analyses, focusing on human
involved distance based interventions [@milner2015] and autonomous
interventions [@torok2020] (see Supplement A for the final strings).

Once search strings were established, the first one hundred search
results of Web of Science were examined together by the authors J.S. and
K.R., establishing a common degree of understanding. After which, both
authors independently searched Web of Science, Scopus and Pubmed;
Systematic searches were last updated on the ...... . Cohen's κ between
both authors was calculated with 0.806. [INTERNAL: As of 30.09 in WOS no
new entries]

## Inclusion and exclusion criteria

All peer reviewed randomized control trial studies were included, which
investigated any form of distance based programs aimed at preventing
self-harming thoughts and/or behaviours, such as suicidal ideation,
suicidal planning, suicide attempts, non-suicidal self-injury (NSSI),
and suicide. Face-to-face meetings were allowed, if these were not part
of the intervention - i.e. for informing, testing or screening purposes.

Interventions which aimed at psychiatric disorders as a primary outcome
and combined outcome measures, such as the total score of SBQ-R, summing
thoughts and behaviours in a total score were excluded. Further, blended
programs, such as phone-calls including home visits were excluded.

## Data Extraction and Coding

Data was coded independently by two authors (J.S and K.R.), if possible,
coding non-imputed results, coded were: Author, Year, control group of
study, Country of Study, Sample type, sample size, intervention type,
sex ratio, Mean Age, the name of the outcome (e.g. self harm), the
duration in Weeks, the Participant Attrition rate, the Follow up Time,
Standard Mean Difference (*SMD)* and variance of *SMD*. In addition, all
outcomes were divided into subgroups according to their characteristics
(see Table 1). Coding sheet is available in the Appendix

+----------+-----------------------------------------------------------+
| M        | Outcome Allocation                                        |
| oderator |                                                           |
| group    |                                                           |
+==========+===========================================================+
| Acts     | Suicide, suicide attempts, deliberate Self Harm.          |
+----------+-----------------------------------------------------------+
| Thought  | Suicidal thoughts, suicidal ideation, suicide planning.   |
+----------+-----------------------------------------------------------+
| Human    | Phone calls, cognitive behavioural treatment,             |
| Involved | personalized letters or personalized e-mails.             |
+----------+-----------------------------------------------------------+
| Au       | Applications, websites, non-individualized letters or     |
| tonomous | non-individualized e-mails                                |
+----------+-----------------------------------------------------------+

: Table 1. Outcome allocation to subgroups.

Once authors finished coding, both combined the sheets, discussing
differences and re-coding affected studies until a unanimous result was
achieved.

## Risk of bias and publication bias

Risk of bias was assessed using the RoB-2 [@sterne2019]. Given emerged
data Trim and Fill was used as the most appropriate publication bias
detection method [@fernandez-castilla2021].

## Statistical Method

To incorporate all outcomes of interest we used a three level
meta-analysis[@vandennoortgate2015][@cheung2019], with Robust variance
estimation (RVE) correction[@hedges2010][@moeyaert2017]. RVE return
valid confidence intervals in presence of dependent data [@park2019a],
while the three level model allowed for outcome-level heterogeneity
investigation [@vandennoortgate2013]. Heterogeneity was estimated using
*Q* Statistics. Models were fitted with restricted maximum likelihood
estimation (REML), RVE correction was based on the correction by
Pustejovsky & Tipton [-@pustejovsky2016].

Calculations were done in R [@base] using the package metafor for the
three level model [@metafor] and the package clubSandwich
[@clubSandwich] for the RVE correction. The R-Code is available in the
Appendix

## Sensitivity Analysis

We employed two sensitivity analyses: (a) Including Non Suicidal Self
Injury (NSSI) as an outcome, (b) Excluding suicide cases as an outcome;
as NSSI [@americanpsychiatricassociation2013a] and suicide cases
[@joiner2005] maybe qualitatively different from other self harming
behaviour.

## Deviation from Pre-Registered Report

Firstly, as the number of included studies was lower than expected the
originally planned method risked underpowered results. We therefore
deviated from the protocol by instead combining RVE correction with a
three-level model, following a current proposition [@pustejovsky2020].

Secondly, we initially wanted to base the publication bias detection
method on the paper by Renkewitz [-@renkewitz2019], but deviated from
the preregistration protocol, as a more accurate paper was published
afterwards [@fernandez-castilla2021]; note that both papers recommend
Trim and Fill, given emerged data.

# Results

```{r include = FALSE}
source("04_Code.R", local = knitr::knit_global())

```

## Emerged Data

We identified 3,287 papers in the databases and an additional 11 studies
from other sources (see Flow Chart), including `r max(Total$Study_ID)`
independent studies. The difference between included studies and
included independent studies is explained by a number of follow up
studies that used the same sample as their parent study. These follow up
studies were statistically treated as dependent to their parent. But for
reporting purposes they were treated as independent studies and as such
also reported within the flow chart and references. Both authors (J.S &
K.R) agreed to 89% about the inclusion and exclusion of studies.

INSERT FIGURE 1. Flow Chart

Overall, `r max(Total$Outcome_ID)` outcomes emerged, with
`r sum(Total$Thinking)` outcomes of the category thinking, predominantly
ideation and planning, and
`r max(Total$Outcome_ID) - sum(Total$Thinking)` outcomes of the category
acting, predominantly deliberate self harm.
`r max(Total$Outcome_ID) - sum(Total$Autonomous)` outcomes evaluated
human involved programs and `r sum(Total$Autonomous)` outcomes evaluated
autonomous programs. The median exposure time of all interventions was
`` r op(median(Total$`Duration (weeks)`)) `` weeks, with a range of
`` r min(Total$`Duration (weeks)`) `` to
`` r max(Total$`Duration (weeks)`) `` weeks. The median time between
post-intervention measures and follow up measures was
`` r median(Total$`Follow_up Time`, na.rm = TRUE) `` weeks, with a range
of `` r max(Total$`Follow_up Time`,  na.rm = TRUE) `` weeks to
`` r min(Total$`Follow_up Time`,  na.rm = TRUE) `` weeks.

The median attrition rate was
`` r op(median(Total$`Total Attrition (%)`, na.rm= TRUE)) `` % with a
maximum of `` r max(Total$`Total Attrition (%)`, na.rm=TRUE) `` % and a
minimum of 0 %. Due to the very different frequencies and types of
interventions quantitative reporting of utilisation was not seen as
sensible.

### Sample Characteristics

In total the sample consisted of n = `r op0(sum(n_mean_perStudy$x))`,
with n= `r sum(AggPost$n)` at post-intervention and n=
`r sum(AggFollow$n)` at follow up, these were to
`` r op(mean(Total$`Sex_(Female)`)) `` % female and on average
`` r op(mean(Total$`Age_(mean)`, na.rm = TRUE)) `` years old. The
youngest mean sample age was
`` r min(Total$`Age_(mean)`, na.rm = TRUE) `` years, the oldest mean
sample age was `` r max(Total$`Age_(mean)`, na.rm = TRUE) `` years.

Most studies were from the United States (*k=* 10), followed by
Australia (*k* = 8); a total of *k =* 6 studies originated from non
westernised educated industrialised democracies (WEIRD).

## Main Analysis

Distance based interventions were effective against suicidal thoughts
and behaviours, at SMD = `r op(OverallRVECI$beta)`
CI95%[`r op(OverallRVECI$CI_L)`; `r op(OverallRVECI$CI_U)`];
Heterogeneity was significant at *Q* (df = `r Overall$dfs`)=
`r op(Overall$QE)`, *p =* `r op(Overall$QEp)`.

The moderator thoughts vs. behaviours was significant, as effectiveness
of distance based interventions against suicidal acts were minimal with
*SMD=* `r op(Overall.TRVECI$beta[1])`
CI95%[`r op(Overall.TRVECI$CI_L[1])`; `r op(Overall.TRVECI$CI_U[1])`],
while interventions aimed at suicidal thinking were more effective
*SMD=* `r op(Overall.TRVECI$beta[2])`
CI95%[`r op(Overall.TRVECI$CI_L[2])`; `r op(Overall.TRVECI$CI_U[2])`],
giving an averaged effectiveness against suicidal thoughts of *SMD =*
`r op(Overall.TRVECI$beta[1] + Overall.TRVECI$beta[2])`; Heterogeneity
became insignificant with *Q* (df = `r Overall.T$dfs`) =
`r op(Overall.T$QE)`, *p =* `r op(Overall.T$QEp)`.

For Controlgroup comparisons two groups could be compared: TAU versus
waitlist and attention placebo controlgroups; results of analysis
differentiating waitlist and attention placebo controlgroups returned
statistically unreliable results. Comparing both groups, waitlist and
attention placebo controlgroups returned significantly stronger effects
with *SMD=* `r op(Overall.CRVECI$beta[1])`
CI95%[`r op(Overall.CRVECI$CI_L[1])`; `r op(Overall.CRVECI$CI_U[1])`],
compared to these studies using a TAU controlgroup were around *SMD=*
`r op(Overall.CRVECI$beta[2])` CI95%[`r op(Overall.CRVECI$CI_L[2])`;
`r op(Overall.CRVECI$CI_U[2])`], less effective; heterogeneity remained
significant = *Q* (df = `r OverallC$dfs`) = `r op(OverallC$QE)`, *p =*
`r op(OverallC$QEp)`.

Graphically inspecting both significant moderators revealed that
suicidal acts and suicidal thoughts were unevenly distributed between
studies using different controlgroups (see figure 2). An exploratory
analysis including both moderators revealed that Controlgroups were no
longer significant, when accounting for suicidal acts and suicidal
thoughts, with *SMD=* `r op(Overall.TRVECITT$beta[3])`
CI95%[`r op(Overall.TRVECITT$CI_L[3])`;
`r op(Overall.TRVECITT$CI_U[3])`]. While suicidal acts and suicidal
thoughts remained a significant moderator for the effectiveness of
interventions, with *SMD =* `r op(Overall.TRVECITT$beta[2])`
CI95%[`r op(Overall.TRVECITT$CI_L[2])`;
`r op(Overall.TRVECITT$CI_U[2])`]; heterogeneity was insignificant = *Q*
(df = `r Overall.TT$dfs`) = `r op(Overall.TT$QE)`, *p =*
`r op(Overall.TT$QEp)`.

INSERT: FIGURE 2. Funnel Plot

Effectiveness of distance based interventions decreased insignificantly
between Post Intervention and Follow up Measurements with *SMD =*
`r op(Overall.FCRVECI$beta[2])` CI95%[`r op(Overall.FCRVECI$CI_L[2])`;
`r op(Overall.FCRVECI$CI_U[2])`]; heterogeneity was significant = *Q*
(df = `r OverallF$dfs`) = `r op(OverallF$QE)`, *p =*
`r op(OverallF$QEp)`.

Human involvement had an insignificant negative impact on effectiveness
*SMD =* `r op(Overall.IRVECI$beta[2])`
CI95%[`r op(Overall.IRVECI$CI_L[2])`; `r op(Overall.IRVECI$CI_U[2])`];
heterogeneity was significant = *Q* (df = `r Overall.I$dfs`) =
`r op(Overall.I$QE)`, *p =* `r op(Overall.I$QEp)`.

## Sensitivity Analysis

Including NSSI in the moderator analysis of thinking vs. behaviour
negligibly impacted the results *SMD* = `r op(Overall.TRVE2$beta[1])` ,
compared to the analysis without NSSI included: *SMD =*
`r op(Overall.TRVE$beta[1])`. Excluding suicides lead to a negligible
reduction in effectiveness as well, comparing effectiveness without
suicides at *SMD* = `r op(Overall.TSRVE$beta[1])`, to the analysis
including suicide cases *SMD =* `r op(Overall.TRVE$beta[1])`.

## Publication Bias and Risk of Bias Assessment

Risk of bias of all *independent* studies was mixed (see Figure 3.).
Using Trim and Fill, no publication bias could be observed.

INSERT: FIGURE 3. Rob-2 Study Quality

# Discussion

We investigated the effectiveness of distance based interventions in
reducing suicidal thoughts and behaviour. On average distance based
interventions reduced suicidal thoughts and behaviours. The Quality of
evidence was good, given a considerable amount of RCT and a notable
number of high and medium quality studies.

### Contextualising results with other Meta-Analyses

Distance based interventions against suicidal behaviours and
face-to-face therapeutic interventions against self harming behaviour
and suicide attempts [@ougrin2015] were comparatively effective. While
the mean score of psychoanalytical face-to-face interventions
[@briggs2019] was three to six times higher, it remained statistically
insignificant. That interventions can be three to six times more
effective than distance based interventions and remain statistically
insignificant shows how low effectiveness against suicidal behaviours
were.

For suicidal thoughts distance based interventions and face-to-face
interventions against suicidal ideation[@bornheimer2020] were again
comparatively effective. Although the generalizability of the
face-to-face meta-analysis is limited, as only psychotic patients were
included in the analysis.

Based on these results Distance based interventions fit perfectly into
the role of low cost, low barrier treatments against suicidal thoughts.
Additionally, as autonomous programs were as effective as human involved
programs, including internet delivered cognitive behavioural treatments
(iCBT) [@buscher2020c], autonomous distance based interventions could
provide treatment opportunities to all 138 Million individuals
experiencing suicidal ideation.

### Research recommendations

Development should be focused on autonomous distance based interventions
that provide low cost community level support against suicidal thoughts.
These programs should include severity measurements, which give a strong
recommendation to seek face-to-face treatment, if the user experiences
suicidal behaviours, including self harm.

### Implications for clinical practice

Distance based interventions give psycho-social service providers a
treatment that they can offer instantly, thereby reducing the number of
delayed or aborted treatments. Given that all interventions showed
higher effectiveness against suicidal thoughts than suicidal acts,
including face-to-face treatment, offering quick and early access to
help is crucial. Furthermore some who would otherwise perceive their
need as insufficient to seek treatment [@bruffaerts2011], might be
convinced to use autonomous intervention, given the lower personal cost
and effort. Concluding: (Autonomous) distance based interventions are
clearly better than no treatment and therefore should always be seen as
a treatment option for an early low intensity intervention.

## Limitations

Given the comprehensive approach of this analysis some limitations must
be noted.

First, heterogeneity was high, but the moderator thinking vs. acting
explained a substantial degree of heterogeneity and, given the large
number of different terms, scales, countries and populations subsumed
within a single analysis, the heterogeneity was surprisingly low.

Secondly, given the power demanding model and the total number of
emerged studies, results might be underpowered, which potentially leads
to too narrow confidence intervals [@tanner-smith2014]. Nonetheless we
observed no indications for notably lacking power in any analysis that
was presented.

nocite:\|
@amadeo2015b,@batterham2018a,@beautrais2010,@bertolote2010,@bush2017,@carter2005,@carter2007,@carter2013,@cedereke2002,@christensen2013a,@comtois2019,@dejaegere2019,@evans2005,@evans1999,@franklin2016,@guille2015,@hassanian-moghaddam2011,@hassanian-moghaddam2017a,@hetrick2017,@hill2019,@hooley2018,@kasckow2016,@luxton2020,@motto2001,@mouaffak2015,@mousavi2014,@niederkrotenthaler2020a,@oconnor2017a,@robinson2012,@rodante2020c,@tighe2017,@vaiva2018,@vaiva2006,@vanspijker2014,@vanspijker2018,@wilks2018a,@wilksch2019,@wei2013

# References

::: {#refs}
:::

# Appendix

## Online Supplement 1

### Search Strings:

Note: Search string was adapted in Scopus adding: TITLE-ABS-KEY. This is
implied when using WOS and Pubmed. As in WOS and Pubmed Title, Abstract
and Keywords were also searched, both strings are in practice identical.

#### For WOS and Pubmed

(("Suic\*" OR "Suicide prevention" OR "self harm\*" OR "self
poisoning\*" OR "self injur\*" OR "self mutilation" ) AND
("telehealth"OR"postcard\*"OR "onlin\*"OR"Online Intervention" OR
"Online Prevent\*" OR "E Intervention" OR "E-Intervention" OR "E
Prevention" OR "Electronic Intervention" OR "Electronic Prevention" OR
"Mobile Intervention" OR "Mobile Prevention" OR "Web-Based\*" OR "Web
Based\*"  OR "Online Support" OR "E Therapy" OR "e-mail\*" OR "e mail\*"
OR "App" OR "Apps" OR "App-Assis\*" OR "mobile-App" OR "mobile health
intervention" OR "telephone" OR "phone based" OR "letter\*") AND (RCT OR
Random\*))

#### Scopus:

( TITLE-ABS-KEY ( ( "Suic\*"  OR  "Suicide prevention"  OR  "self
harm\*"  OR  "self poisoning\*"  OR  "self injur\*"  OR  "self
mutilation" )  AND  ( "telehealth"  OR  "postcard\*"  OR  "onlin\*"  OR 
"Online Intervention"  OR  "Online Prevent\*"  OR  "E Intervention"  OR 
"E-Intervention"  OR  "E Prevention"  OR  "Electronic Intervention"  OR 
"Electronic Prevention"  OR  "Mobile Intervention"  OR  "Mobile
Prevention"  OR  "Web-Based\*"  OR  "Web Based\*"  OR  "Online Support" 
OR  "E Therapy"  OR  "e-mail\*"  OR  "e mail\*"  OR  "App"  OR  "Apps" 
OR  "App-Assis\*"  OR  "mobile-App"  OR  "mobile health intervention" 
OR  "telephone"  OR  "phone based"  OR  "letter\*" )  AND  ( rct  OR 
random\* ) ) )

## Online Supplement 2

### Code.R

Coding_Sheet.csv
